
# Day-0 : Installation & Environment Setup (Linux) {-}

## Learning Objectives
- Install R (version 4.5)
- Install RStudio
- Verify a working R environment

---

## Installing R (Ubuntu/Debian)

Run the following commands in a terminal:

```{r, engine='bash', eval=FALSE}

sudo apt update

sudo apt install -y software-properties-common dirmngr

sudo wget -qO- https://cloud.r-project.org/bin/linux/ubuntu/marutter_pubkey.asc \
  | sudo tee /etc/apt/trusted.gpg.d/cran_ubuntu_key.asc

sudo add-apt-repository "deb https://cloud.r-project.org/bin/linux/ubuntu jammy-cran40/"

sudo apt update
sudo apt install -y r-base=4.5* r-base-dev=4.5*
```

## You can also follow below link for installations

- https://cran.r-project.org/bin/linux/ubuntu/

## Installing RStudio Desktop (Linux)

- RStudio provides a user-friendly interface for working with R.

- Step 1: Download RStudio

- Go to the official RStudio download page:

- https://posit.co/download/rstudio-desktop/

- Download the Linux → Debian/Ubuntu (.deb) installer.

### Step 2: Install RStudio

- Navigate to the downloaded file and run:

```{r, engine='bash', eval=FALSE}

sudo dpkg -i rstudio-*.deb
sudo apt --fix-broken install

```

### Step 3: Launch RStudio

- You can start RStudio by:

- Searching “RStudio” in the application menu

- Or running in terminal:

```{r, engine='bash', eval=FALSE}
rstudio

```


