
# Day-1 : Fundamentals of R Programming 

## Learning Objectives

By the end of this day, students will be able to:

- Understand what R is and where it is used
- Use R as a calculator
- Create and manipulate basic R objects
- Work with vectors and data frames
- Use basic R functions and help documentation


## What Is R?

R is a **programming language and environment** designed for:

- Statistical analysis
- Data visualization
- Reproducible research

R is widely used in:
- Statistics
- Epidemiology
- Genomics
- Social sciences
- Machine learning research

One of R’s biggest strengths is its **ecosystem of packages**, including  
**CRAN** and **Bioconductor**.

---

## RStudio Interface (Quick Tour)

RStudio provides a user-friendly interface for working with R.

Main components:
- **Console** – where commands are executed
- **Script Editor** – where you write and save code
- **Environment** – shows created objects
- **Plots / Files / Help** – outputs and documentation

> 💡 Tip:  
> Always write code in a **script file**, not directly in the Console.

---

## Using R as a Calculator

Let’s start with simple arithmetic.

```{r}
2 + 3
10 - 4
5 * 6
8 / 2
```


Additional useful operators:

```{r}
2^3 # Exponentiation
10 %% 3 # Remainder
10 %/% 3 # Integer division
```

## Creating Objects in R

R stores values in objects.

Assignment is typically done using <-:

```{r}
x <- 10
y <- 3


```

Use the objects in calculations:

```{r}
x + y
x * y
```

💡 **Tip**
The "=" operator also works, but <- is the recommended R convention.

---

## Basic Data Types

Common data types in R include:

```{r}
numeric_value <- 3.14
integer_value <- 10L
character_value <- "genomics"
logical_value <- TRUE

```

Check the type of an object:
```{r}

class(numeric_value)
class(character_value)
class(logical_value)

```

---

## Data Structures

### Vector

A vector is a collection of elements of the same data type.

```{r}

numbers <- c(1, 2, 3, 4, 5)
genes <- c("TP53", "BRCA1", "EGFR")
flags <- c(TRUE, FALSE, TRUE)

```

Indexing vectors:

```{r}

numbers[1]
numbers[2:4]

```

## Vectorization: Thinking in Vectors

One of the key ideas in R is **vectorization**.

In R, most operations are designed to work on **entire vectors at once**, rather than
one element at a time. This makes R code:

- Simpler to write
- Easier to read
- Faster to execute

This is different from many other programming languages, where you often need loops
to process each value individually.

```{r}

numbers * 2
numbers + 10

```


### Data Frames

A data frame is a table-like structure where each column can be a different data type.

```{r}
df <- data.frame(
id = 1:5,
age = c(23, 35, 41, 29, 50),
group = c("A", "B", "A", "B", "A")
)

df

```

Inspecting a data frame:

```{r}
str(df)
summary(df)
head(df)

```

Accessing columns:

```{r}

df$age
df[, "group"]

```


## Getting Help in R

R has built-in help documentation.

```{r}
?mean
help(mean)

help.search("regression") # gives difference functions to perform regression using in-built functions or downloaded packages

```


## Hands-on Exercises

### Exercise 1

Create a numeric vector of 10 values and calculate:

- Mean

- Median

- Standard deviation

### Exercise 2

Create a data frame with:

- Name

- Age

- Field of study

- Display its structure and summary.

### Exercise 3

Explore the built-in iris dataset:

- View the first 6 rows

- Find the mean of Sepal.Length




