# Day-3 : Statistical Analysis Using R

---

## Learning Objectives

By the end of this day, students will be able to:

- Understand descriptive and inferential statistics
- Summarize data numerically and graphically
- Understand the role of probability in statistics
- Perform common hypothesis tests in R
- Interpret p-values and confidence intervals
- Fit and interpret basic statistical models

---

## Types of Data

Understanding data types is essential for choosing the right statistical test. Also these are default data type in many statistical packages.

- **Numeric** – continuous or discrete values  
- **Categorical** – groups or labels  
- **Binary** – two outcomes (e.g., yes/no)


### Why Do We Need Statistics? 

In biology, data are variable by nature. Even under controlled conditions, measurements differ due to biological and technical variability.
Statistics provides tools to summarize data, quantify uncertainty, and draw conclusions from samples.

Statistics helps us to:

- Summarize complex data
- Understand natural variation
- Make decisions using samples instead of entire populations

Key ideas:

- Data ≠ truth
- Samples ≠ populations
- Variation is expected and normal

Statistics gives us tools to decide whether observed differences are meaningful or simply due to random variation.

Example: Two treatments show different average gene expression levels — is the difference real or random? Stats will help in answering that question.

---

## Descriptive vs Inferential Statistics

Statistics can be broadly divided into two categories:

- **Descriptive statistics**  
  Summarize and describe the observed data.

- **Inferential statistics**  
  Use sample data to draw conclusions about a larger population.

Examples:

- Calculating the mean sepal length → *descriptive*
- Testing whether two species differ in mean sepal length → *inferential*

In this session, we start with **descriptive statistics** and then move to **inferential statistics**.

---

## Descriptive Statistics

Descriptive statistics summarize the main characteristics of a dataset.

### Measures of Central Tendency

Measures of central tendency describe the “typical” value of a dataset.

- **Mean** – arithmetic average (sensitive to outliers)
- **Median** – middle value (robust to outliers)

Measures of Variation (Spread):

Variation describes how spread out the data are.

Variance – average squared deviation from the mean

Standard deviation (SD) – typical distance from the mean

range - difference in maximum and minimum value

Inter-quartile range: difference in 3rd and 1st quintile


### Numeric Summaries

```{r}

mean(iris$Sepal.Length)
median(iris$Sepal.Length)
sd(iris$Sepal.Length)
range(iris$Sepal.Length)

```

### Grouped summaries:

```{r}
library(dplyr)

iris %>%
group_by(Species) %>%
summarise(
mean_sepal = mean(Sepal.Length),
sd_sepal = sd(Sepal.Length)
)

```


### Visualizing Distributions

Histogram

```{r}
library(ggplot2)

ggplot(iris, aes(x = Sepal.Length)) +
geom_histogram(bins = 20)

```

Boxplot

```{r}
ggplot(iris, aes(x = Species, y = Sepal.Length)) +
geom_boxplot()

```

### From Description to Inference

Inferential statistics help us answer questions such as:

Are observed differences real?

Or could they be explained by natural variation?

To answer these questions, we use hypothesis testing.


### Populations vs Samples

Population: The complete set of possible outcomes (e.g., all students in the class).

Sample: A subset of the population (e.g., a random group of students measured for some trait).

In practice, we usually work with samples, because measuring entire populations is often impossible.


Suppose you have 50 students, and you want to predict the gender distribution. You know historically your class has roughly equal number of males and females


```{r}
set.seed(123)

# Simulate the population of 50 students (M = male, F = female)

students <- sample(c("M", "F"), size = 50, replace = TRUE)
table(students)

# Notice that the number of males and females in your sample can vary just by chance.

```

### Sampling Distribution

What if you take random samples of 10 students from your class multiple times?
Each sample will have a different number of males and females — this is sampling variation.


```{r}

set.seed(123)
sample_gender_counts <- replicate(1000, {
sample10 <- sample(students, 10)
sum(sample10 == "M")  # count number of males in each sample
})

# Visualize

library(ggplot2)

ggplot(data.frame(male_count = sample_gender_counts),
       aes(x = male_count)) +
  geom_histogram(
    breaks = 0:10,
    fill = "lightblue",
    color = "black"
  ) +
  geom_vline(
    xintercept = mean(sample_gender_counts),
    color = "red",
    linewidth = 1
  ) +
  labs(
    title = "Distribution of Male Students in Random Samples of 10",
    x = "Number of Males",
    y = "Frequency"
  ) +
  theme_minimal()

```

Each bar shows how likely a particular number of males occurs in a random sample of 10 students.

The red line shows the average number of males across all samples.

This histogram is a sampling distribution, the foundation of statistical inference.

### From Variation to Statistical Testing

Observed differences do not automatically imply a real or meaningful effect.

Statistics helps us answer questions like:

Is the observed difference larger than what we would expect by chance?

Is our sample consistent with a known or assumed population value?

Example: Testing a Population Proportion

We can formally test whether the proportion of males in the class differs from 50%.

```{r}
prop.test(sum(students == "M"),
length(students),
p = 0.5)

# The test output includes a p-value, which helps us assess whether the observed proportion is consistent with the assumed population proportion.
```

### Interpreting p-values (Common Mistakes)

A p-value is NOT:
- The probability that the null hypothesis is true
- A measure of effect size

A p-value IS:
- A measure of how surprising the data are if H₀ were true

### Confidence Interval Interpretation 

A 95% confidence interval means that if we repeated the experiment many times, about 95% of the intervals would contain the true value.

## Introduction to Hypothesis Testing

Hypothesis testing uses probability to evaluate whether observed differences
are likely due to random variation or represent a real effect.


### Key concepts:

- Null hypothesis (H₀): no effect or no difference

- Alternative hypothesis (H₁): effect or difference exists

- p-value: probability of observing the data or statistics assuming H₀ is true

### t-test

One-Sample t-test

Tests whether the mean differs from a known value.

```{r}

t.test(iris$Sepal.Length, mu = 5)

```

### Two-Sample t-test

Compare Sepal Length between two species.

```{r}

t.test(
Sepal.Length ~ Species,
data = subset(iris, Species %in% c("setosa", "versicolor"))
)

```

### Non-Parametric Tests (Brief Overview)

When data do not meet assumptions of parametric tests
(e.g., normality), non-parametric tests are used.

Example: Wilcoxon rank-sum test (shown for reference)

```{r}

wilcox.test(
  Sepal.Length ~ Species,
  data = subset(iris, Species %in% c("setosa", "versicolor"))
)

```

### Analysis of Variance (ANOVA)

ANOVA compares means across more than two groups.

```{r}
anova_model <- aov(Sepal.Length ~ Species, data = iris)
summary(anova_model)

```


### Correlation Analysis

measures the strength and direction of the linear relationship between two variables. varies from -1 to 1

```{r}

cor(iris$Sepal.Length, iris$Petal.Length)

```

### Linear Regression

Linear regression models the relationship between variables.

```{r}
lm_model <- lm(Sepal.Length ~ Petal.Length, data = iris)
summary(lm_model)

```

### Visualizing the Regression

```{r}
ggplot(iris, aes(x = Petal.Length, y = Sepal.Length)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE)

```


### Interpreting Statistical Results

When reporting results, always consider:

- Effect size

- Confidence intervals

- p-values

- Practical or biological relevance

⚠️ Statistical significance does not always imply practical importance.


**Example: Statistical Significance vs Practical Importance**

Scenario:

Suppose we are studying whether a new treatment increases gene expression compared to a control.

Outcome: Gene expression level (continuous)

Two groups: Control vs Treatment

Sample size: 100 per group (large experiment)

```{r}
set.seed(42)

n <- 100

control <- rnorm(n, mean = 10, sd = 2)
treatment <- rnorm(n, mean = 10.3, sd = 2)  # small increase

data <- data.frame(
  expression = c(control, treatment),
  group = rep(c("Control", "Treatment"), each = n)
)

# Descriptive Statistics

data %>%
  group_by(group) %>%
  summarise(
    mean_expression = mean(expression),
    sd_expression = sd(expression)
  )

# Interpretation:
# 
# - Treatment has a slightly higher mean expression
# 
# - The difference looks small relative to variability

# Visual Comparison
library(ggplot2)

ggplot(data, aes(x = group, y = expression)) +
  geom_boxplot(fill = "lightblue") +
  labs(
    title = "Gene Expression by Treatment Group",
    y = "Expression Level"
  ) +
  theme_minimal()

# Inferential Statistics: Two-Sample t-test

test <- t.test(expression ~ group, data = data)
test

# What question is the test answering?
# 
# Is the observed difference in mean expression between Control and Treatment larger than what we’d expect by random chance?

# Test statistic (this is the key number)
# t = -0.21754
# 
# The t-statistic measures:
# 
# How far apart the two group means are
# relative to the variability in the data
# In plain words:
# Difference between means ÷ random noise

# Signal = difference between means
# Noise = variability within groups

# signal << noise  →  t ≈ 0  →  large p-value

# Large |t| → strong evidence of a real difference
# 
# Small |t| → difference is tiny compared to variation
# Here:
# 
# t = -0.22 → very close to zero
# 
# Means are almost the same compared to variability
# 
# (The negative sign just tells you direction, not importance.)
# 
# What does this mean?
# 
# If there were no real difference between groups,
# there is an 82.8% chance of observing a difference at least this large just by random sampling.
# 
# ️ That is very high
# ️ So the observed difference is not unusual at all
# 
# Conclusion: No statistical evidence of a difference.
# 
# Confidence interval
# 95 percent confidence interval:
#  -0.6039960  0.4839901

# This interval represents:
# 
# Plausible values for the true difference in means
# (Control − Treatment)
# Key observation:
# 
# The interval includes 0
# 
# ️ Zero difference is completely plausible
#  Supports the non-significant p-value

test$conf.int

# Interpretation
# We are 95% confident the true mean difference lies in this interval
# The interval does include zero, so the difference is statistically in-significant

# Effect Size (Cohen’s d)

mean_diff <- mean(treatment) - mean(control)
pooled_sd <- sqrt((sd(control)^2 + sd(treatment)^2) / 2)

cohens_d <- mean_diff / pooled_sd
cohens_d


```

## t-Statistic vs Effect Size (Cohen’s d): The Math

Understanding the mathematical difference between the **t-statistic** and **effect size**
helps clarify why statistical significance and practical importance are not the same.

---

### 1️⃣ Two-Sample t-Statistic (Welch t-test)

For two groups with sample means \( \bar{x}_1 \) and \( \bar{x}_2 \):

\[
t =
\frac{\bar{x}_1 - \bar{x}_2}
{\sqrt{\frac{s_1^2}{n_1} + \frac{s_2^2}{n_2}}}
\]

Where:

- \( \bar{x}_1 - \bar{x}_2 \) = difference in sample means  
- \( s_1^2, s_2^2 \) = sample variances  
- \( n_1, n_2 \) = sample sizes  

🔑 **Key point**

The denominator contains \( n_1 \) and \( n_2 \).  
As sample size increases, the denominator becomes smaller, making the
t-statistic **larger even if the mean difference stays the same**.

This is why large datasets can produce very small p-values for very small effects.

---

### 2️⃣ Effect Size: Cohen’s d

Cohen’s \( d \) standardizes the mean difference using **variability only**:

\[
d =
\frac{\bar{x}_1 - \bar{x}_2}{s_{\text{pooled}}}
\]

Where the pooled standard deviation is:

\[
s_{\text{pooled}} =
\sqrt{
\frac{(n_1 - 1)s_1^2 + (n_2 - 1)s_2^2}
{n_1 + n_2 - 2}
}
\]

Where:

- \( s_{\text{pooled}} \) = pooled standard deviation  
- \( n_1 - 1 \), \( n_2 - 1 \) = degrees-of-freedom weights  

🔑 **Key point**

Cohen’s \( d \):

- Uses sample size **only to weight variances**
- Does **not** shrink with increasing \( n \)
- Measures the **magnitude of the effect**, not the strength of evidence

---

### Important Takeaway

> A result can be statistically significant (large t, small p-value)  
> but have a small effect size (small Cohen’s d).


## Machine Learning (Statistical Learning): A Gentle Introduction

In this course, **machine learning** is introduced as an extension of **statistical modeling**, with a different primary goal.

While traditional statistical analysis often focuses on:
- Estimating effects
- Hypothesis testing
- Interpretation

Machine learning (also called **statistical learning**) focuses on:
- Prediction
- Generalization to new data
- Model performance on unseen observations

---

### Statistics vs Machine Learning (Conceptual View)

| Aspect | Statistical Modeling | Machine Learning |
|------|----------------------|------------------|
| Primary goal | Inference & explanation | Prediction |
| Key questions | “Is this effect real?” | “How well does this predict?” |
| Outputs | Coefficients, p-values | Predictions, errors |
| Evaluation | p-values, confidence intervals | Prediction error |

Both approaches use **the same data**, and often **the same models**,
but answer different questions.

---

### A Simple Example: Linear Regression as Machine Learning

Linear regression can be used not only for inference, but also for prediction.

```{r}
set.seed(123)

# Simulated data
x <- rnorm(100)
y <- 2 * x + rnorm(100)

# Train-test split
train_id <- sample(seq_along(x), 70)

x_train <- x[train_id]
y_train <- y[train_id]
x_test  <- x[-train_id]
y_test  <- y[-train_id]

# Fit model on training data
model <- lm(y_train ~ x_train)

# Predict on new data
predictions <- predict(model,
                       newdata = data.frame(x_train = x_test))

head(predictions)

```

Here:

- The model is trained on one part of the data
- Predictions are made on new, unseen data

### Evaluating Predictions

Instead of p-values, we evaluate prediction error.

```{r}

mean((predictions - y_test)^2)

```

### Types of Machine Learning Tasks

Most machine learning problems fall into two broad categories:
**regression** and **classification**.

---

### Regression Tasks

Regression is used when the outcome variable is **continuous**.

Examples:

- Predicting blood pressure
- Predicting gene expression levels
- Predicting healthcare costs

Typical evaluation metrics for regression include:

- **Mean Squared Error (MSE)** – average squared prediction error
- **Root Mean Squared Error (RMSE)** – error on the original scale
- **Mean Absolute Error (MAE)** – average absolute difference
- **R-squared** – proportion of variance explained

In this course, we have seen regression using **linear models**.

---

### Classification Tasks

Classification is used when the outcome variable is **categorical**.

Examples:
- Disease vs healthy
- Treatment responder vs non-responder
- Tumor type classification

Common evaluation metrics for classification include:

- **Accuracy** – proportion of correct predictions
- **Sensitivity (Recall)** – ability to detect positive cases
- **Specificity** – ability to detect negative cases
- **Precision** – proportion of true positives among predicted positives

The choice of metric depends on the **problem context**, especially in biomedical applications.

**High-Dimensional Classification Using caret**

In machine learning, we typically follow a **standard workflow**:
 
1. Split data into training and test sets  
2. Train a model on training data  
3. Evaluate performance on unseen test data  
 
The `caret` package provides a unified interface for this workflow.


```{r}

### Simulated High-Dimensional Dataset

# We simulate a gene expression–like dataset:
# 
# - 60 samples
# - 100 features (genes)
# - Binary outcome (Disease vs Healthy)

set.seed(123)

library(caret)

n_samples  <- 60
n_features <- 100

# Feature matrix
X <- matrix(rnorm(n_samples * n_features),
            nrow = n_samples,
            ncol = n_features)

# Outcome variable
y <- factor(rep(c("Healthy", "Disease"),
                each = n_samples / 2))

# Introduce signal in a few genes
X[y == "Disease", 1:5] <- X[y == "Disease", 1:5] + 1

data_ml <- data.frame(Class = y, X)

dim(data_ml)
table(data_ml$Class)

# train test(validation) split

set.seed(123)

train_index <- createDataPartition(data_ml$Class,
                                   p = 0.7,
                                   list = FALSE)

train_data <- data_ml[train_index, ]
test_data  <- data_ml[-train_index, ]

# Model Training with caret
# We train a logistic regression classifier.

ctrl <- trainControl(
  method = "cv",
  number = 5
)

model_caret <- train(
  Class ~ .,
  data = train_data,
  method = "glm", # rf, svm, gbm, xgboost
  family = "binomial",
  trControl = ctrl,
  metric = "Accuracy"
)

model_caret

pred_class <- predict(model_caret, newdata = test_data)

confusionMatrix(pred_class, test_data$Class)

```

---

### Advanced Machine Learning (Overview)

Beyond linear models, many advanced algorithms are commonly used in practice.

Examples include:

- **Decision trees** – rule-based models
- **Random forests** – ensembles of decision trees
- **Support vector machines (SVMs)** – margin-based classifiers
- **Neural networks** – layered models for complex patterns

These methods are powerful, but require:

- Larger datasets
- Careful tuning
- Proper validation strategies

> 💡 Note:  
> Exploring these algorithms is beyond the scope of this course and
> typically covered in dedicated machine learning classes. For R, you can explore packages like caret and tidymodels specifically for ML tasks.

**Core Concepts You Must Know in Machine Learning**

Before applying advanced machine learning algorithms, it is important to understand how models are trained, evaluated, and controlled.
These ideas apply to all machine learning methods.

**Train–Test Split (Why We Need It)**

- In machine learning, the goal is generalization:
- How well does the model perform on new, unseen data?
- To test this, we split data into:
- Training set – used to fit the model (Learn patterns)
- Test set – used only for final evaluation (Check generalization)

If we evaluate a model on the same data used to train it, we get over-optimistic performance.

**Overfitting vs Underfitting**

Machine learning models can fail in two ways:

- Underfitting:	Model too simple, misses important patterns
- Overfitting:	Model too complex, memorizes noise

intuition:

- A straight line for complex data → underfitting
- A curve passing through every point → overfitting

The goal is good bias–variance tradeoff.

**Cross-Validation (CV)**

Cross-validation is a way to estimate model performance more reliably.

Instead of one train–test split:

- Data is split into K folds
- Model is trained K times
- Each fold is used once as validation data

Most common: 5-fold-cv,  10-fold-cv

Train → Validate → Repeat → Average performance

Why it matters:

- Reduces dependence on one random split
- Gives more stable performance estimates
- Standard practice in ML

**Regularization (Controlling Model Complexity)**

Regularization prevents models from becoming too complex.

Idea:

- Penalize large or unnecessary coefficients

Common types:

- L1 (Lasso) – performs feature selection
- L2 (Ridge) – shrinks coefficients smoothly

Why regularization is critical:

Especially important in high-dimensional data

- Improves generalization
- Reduces overfitting

**Evaluation Metrics Depend on the Task**

Unlike classical statistics, machine learning does not rely on p-values.

Regression:	MSE, RMSE, MAE, R²
Classification:	Accuracy, Sensitivity, Specificity, Precision

Important:

- No single metric fits all problems
- Metric choice depends on scientific or clinical context

Note:
For imbalanced datasets, where one outcome is rare (for example, many cancer or rare disease cases), the real-world class distribution is often not balanced.

In such situations:

Overall accuracy can be misleading

Metrics like sensitivity, specificity, and precision become more important

These metrics help ensure that models support good clinical and biological decision-making, especially when missing rare cases has serious consequences.

In later machine learning courses, students will learn techniques to handle imbalanced data explicitly through resampling, cost-sensitive learning, and threshold tuning.

**Validation Is Not Optional in ML**

In statistics:

 - Focus is often on estimation and hypothesis testing

In machine learning:

- Focus is on out-of-sample performance

This means:

- Always separate training and evaluation
- Always validate models
- Never trust performance on training data

## Hands-on Exercises

### Exercise 1

- Calculate descriptive statistics for Petal.Width grouped by species.

### Exercise 2

- Perform a t-test comparing Petal.Length between:

    setosa & virginica

### Exercise 3

- Fit a linear model predicting Petal.Width from Petal.Length.

### Exercise 4 (Challenge)

- Create a boxplot of Petal.Length by species and interpret the differences.
