# Day-4 : Genomic Data Exploration Using R

## Learning Objectives

By the end of this day, students will be able to:

- Understand common genomic data types
- Learn what Bioconductor is and why it is used
- Work with expression data in R
- Use Bioconductor data structures
- Perform basic exploratory analysis on genomic datasets

---

## What Is Genomic Data?

Genomic data refers to information derived from DNA, RNA, or related molecules.

Common genomic data types include:

- **Gene expression data** (RNA-seq, microarrays)
- **Genomic coordinates** (BED, GFF, GRanges)
- **Variant data** (VCF)
- **Annotation data** (gene names, pathways)

In this course, we focus on **gene expression data**.

---

## What Is Bioconductor?

**Bioconductor** is a collection of R packages designed for:

- Genomic data analysis
- Biological annotation
- Reproducible research

Key features:

- Standardized data structures
- Well-documented packages
- Strong integration with R statistics

Bioconductor complements CRAN.

---

## Installing Bioconductor

Install Bioconductor (only once):

```{r, eval=FALSE}

# install.packages("BiocManager") [unhash it for first run]

# BiocManager::install() [unhash it for first run]

```

Install required packages:

```{r}

# BiocManager::install(c("SummarizedExperiment", "airway", "Biobase")) [unhash it for first run]

```

Loading Required Libraries

```{r}
library(SummarizedExperiment)
library(airway)
library(Biobase)
library(dplyr)
library(ggplot2)

```

Example Dataset: airway

The airway dataset contains RNA-seq gene expression data from an experiment
studying the effect of dexamethasone on human airway smooth muscle cells.

Load the dataset:

```{r}
data(airway)
airway

```

Understanding SummarizedExperiment

SummarizedExperiment is a core Bioconductor data structure.

It contains:

- assay data – expression matrix

- rowData – gene-level information

- colData – sample-level metadata

Exploring the Expression Data
Assay Data (Counts Matrix)

```{r}
assay(airway)[1:5, 1:5]

```

Dimensions:

```{r}
dim(assay(airway))

```

Sample Metadata (colData)

```{r}
colData(airway)

```

Convert to data frame:

```{r}
sample_info <- as.data.frame(colData(airway))
head(sample_info)

```

Gene Metadata (rowData)

```{r}
rowData(airway)

```


Basic Quality Checks
Library Sizes

```{r}
library_sizes <- colSums(assay(airway))
library_sizes

```

Visualize library sizes:

```{r}
qplot(library_sizes, bins = 10) +
labs(
title = "Library Size Distribution",
x = "Total Read Counts",
y = "Number of Samples"
)

```

